A Java Wrapper For Embarrassingly Parallel Problems

### 2013-11-20 paper/paper.tex
Renamed Manipulator and did some work on paper and readme

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that 
can easily be broken up into identital components that do not need to interact 
to find the solution. Several problems in the biological sciences are 
embarrsingly parallel but they are difficult to rewrite to use multithreading, 
because they are either poorly written or too complex. We have written a 
framework, implemented in Java, that will execute the serial program in several 
threads. These threads working together help cut down the total runtime
exponentially. More statistics, savable configurations and more options will be 
implemented next, in addition to better testing and refactored.
\end{abstract}

@@

The framework provides a "wrapper" for the original program that mimics the
behavior of the original program. It calls the serial programs on multiple Java 
threads. Each thread acts like the serial application, reading a subset of the 
input the same and creating the output in the same format. This paper describes
a framework a framework that makes it easy to parallelize an embarrassingly
parallel program that reads from a file, writes to a file and whose input can be
broken up into work units by a regular expression. The framework gives an
exponential decrease in runtime as more processor help in the computation. This
decrease levels off between 50 and 100\% CPU usage due to the other tasks on the
machine that take up memory.

@@

The motivation for this effort was with research into different forms of the 
species \emph{Cicer arietinum}, the chikckpea. The two main forms, Kabuli and 
Desi, are very different in size and color but we mostly wanted to see why the 
Desi is more disease resistance and have more effective root nodule symbiosis 
and why the Kabuli has a large but soft shell. These two types are both 
cultivated and have been under human selection for several thousand years. We 
wanted to use a Ka/Ks equation to quantify the evolution and selection of these 
chickpeas. We found a Ka/Ks program writen by Zhang et al. \cite{kaks} but had 
three variants that we wanted to be analyzed, each with thousands of sequences. 
The Ka/Ks Calculator program takes somewhere from five to thirty seconds to 
process one pair of sequences.

@@

The Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. To deal with large
amounts of data, the best method is often to group them into clusters. 
Hierarchical clustering algorthms take the data and then organize into
hierarchical structure, usually a dendrogram, based on a priority matric. There 
are hierarchical clustering algorithms that can be implemented in parallel using
an embarrassingly parallel approach.  With several applications that would 
benefit from parallelism in this field, it was obvious it would be desirable to 
develop a framework for creating wrapper programs. \cite{cluster}

% TALK ABOUT PAPER HERE. AND MAYBE THE POWERPOINT TOO.

In this version of the framework, several components are being added. First, 
there is a wizard style menu that will allow the end-user to select not only the 
program that wish to use, but also to manage other settings for the run. A new 
IO class splits the input and merges the output more efficiently. Lastly, in 
order to look at how well the framework performs, the threads keep their own 
statistics, which will be used to evaluate the effectiveness of the framework 
and to find potential enhancements and help end-users determine whether their 
use of the framework is efficient.

\section{Proposed Solution}
\begin{figure}
{\resizebox{5.2in}{!}{\includegraphics{figures/uml.png}}}
\caption{UML for the embarrassingly parallel framework}
\label{fig:uml}
\end{figure}
The KaKs Calculator is not the only program to use the framework; there are
various potential programs that the framework could support. The only 
requirements for the this program are that it process the input in an 
embarrassingly parallel manner and that the input comes from a file and the 
processed results are also writen to a file. The framework needs to support 
various kinds of input files and be able to chunk up the work accordingly. The 
Ka/Ks Calculator's input consists of a header and then two protein sequence 
lines followed by a blank line. The framework uses that blank line to know when 
one chunk of work has ended and another has begun. The user can input a regular 
expression to define what lines of input make up what chunks. This occurs in 
the ChunkManager class.

The other purpose of the ChunkManager class is to combine the resulting chunks 
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. The
ChunkManager's second job is to retrieve those results and construct a real
output file. The real output file must look the same as the output file would
look if the program were run serially; the output must come in the same order as
the inputs in the input file and headers may need to be deleted from the
separate files and added to the combined file.

@@

Lastly, the worker threads will keep records during execution. The threads 
records their uptime and the number of pieces of work they execute. The runtime 
for each piece of work is also recorded. This information will be used by us to
evaluate the effectiveness of our approach, but can also provide useful
information the can help a user tine a particular application on a particular
type of input data. The statistics for each thread can be found in the Worker 
subclass and the individual Chunk classes will store their own statstics about 
runtime and size.

@@

To observe how the framework changes the runtime of sequential programs when
using varying numbers of threads, I designed a simple experiment. The Ka/Ks 
Calculator was run through the framework with ten different samples, each 
containing sixteen units of work. Each sample was run by the framework at five 
different level of numbers of worker threads, starting at 1 and increasing by 
powers of two until the 100\% CPU usage (16 processors). These tests were run
with minimal other activity on the server.

@@

There are several next steps in the pipeline. In addition to using the
statistics to try to boost speed, adding savable configurations is at the top of
the list. This will allow users to use the same configuration without having to
go through the wizard if they need to redo a run. More extensive unit testing
will be implemented. The ability to chunk multiple pieces of work may be
implemented in order to avoid potential start up costs in the executable. 
Finally, other pleasing parallel problems will be executed by the wrapper so 
that more potential features will be discovered.

Using Java was a good choice for implementation ease and future implementation
extension by others, but it may not produce the fastest results. Languages
like Hadoop, Clojure or other multithread oriented language may have been better
due to their parallel nature. In addition to a single machine working to process
the data quickly, we can also look to the cloud for help by using a networked 
solution like MPI or a cloud solution like Amazon instances. This would
potentially allow for an even greater time decrease because the number of
machines that could help is most likely larger than the number of processors on
one machine.

\begin{thebibliography}{1}
\bibitem{history}
Moler, Cleve. ``\emph{Matrix Computation on Distributed Memory Multiprocessors."
}, Hypercube Multiprocessors, 1986: Proceedings of the First Conference on
Hypercube Multiprocessors, Knoxville, Tennessee, August 24-27, 1985. By Michael
T. Heath. Philadelphia: SIAM, 1986. N. pag. Print.
\bibitem{kaks}
Zhang Z, Li J, Zhao XQ, Wang J, Wong GK, Yu J., \emph{KaKs Calculator: 
calculating Ka and Ks through model selection and model averaging},
Genomics Proteomics Bioinformatics, November 2006.
\bibitem{cluster}
Rua, Xu, Wunsch, Donald II, \emph{Survey of Clustering Algorithms},
IEEE Transactions on Neural Networks VOL. 16, NO. 3, MAY 2005
\end{thebibliography}

>>>>

\begin{abstract}
An embarrassingly parallel problem, also called pleasingly parallel, is one that 
can easily be broken up into identital components that do not need to interact 
to find the solution. Several problems in the biological sciences are 
embarrsingly parallel but they are difficult to rewrite to use multithreading, 
because they are either poorly written or too complex. We have written a 
framework, implemented in Java, that will execute the serial program in several 
threads. These threads working together help cut down the total runtime
exponentially. The next step is to seperate the framework into a front-end that
will run locally and the back-end which recieves blocks of work and does the
computations. This back-end should be running in the cloud, on a machine who is
dedicated to running embarrassingly parallel tasks only.
\end{abstract}

@@

The framework provides a "wrapper" for the original program that mimics the
behavior of the original program. It calls the serial programs on multiple Java 
threads. Each thread acts like the serial application, reading a subset of the 
input the same and creating the output in the same format. This paper describes
a framework that makes it easy to parallelize an potentially embarrassingly
parallel program that reads from a file, writes to a file and whose input can be
broken up into work units by a regular expression. The framework gives an
exponential decrease in runtime as more processor help in the computation. This
decrease levels off between 50 and 100\% CPU usage due to the other tasks on the
machine that take up memory.

@@

The motivation for this effort was with research into different forms of the 
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and 
Desi, are very different in size and color but we mostly wanted to see why the 
Desi is more disease resistance and have more effective root nodule symbiosis 
and why the Kabuli has a large but soft shell. These two types are both 
cultivated and have been under human selection for several thousand years. We 
wanted to use a Ka/Ks equation to quantify the evolution and selection of these 
chickpeas. We found a Ka/Ks program writen by Zhang et al. \cite{kaks} but had 
three variants that we wanted to be analyzed, each with thousands of sequences. 
The Ka/Ks Calculator program takes somewhere from five to thirty seconds to 
process one pair of sequences.

@@

%TODO: get a better source? Talk about UCLUST and octupus
The Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. To deal with large
amounts of data, the best method is often to group them into clusters. 
Hierarchical clustering algorthms take the data and then organize into
hierarchical structure, usually a dendrogram, based on a priority matric. There 
are hierarchical clustering algorithms that can be implemented in parallel using
an embarrassingly parallel approach.  With several applications that would 
benefit from parallelism in this field, it was obvious it would be desirable to 
develop a framework for creating wrapper programs. \cite{cluster}

%TODO: TALK ABOUT CLOUD PAPER HERE. AND MAYBE THE POWERPOINT TOO.

\section{Proposed Solution}
\begin{figure}
{\resizebox{5.2in}{!}{\includegraphics{figures/uml.png}}}
\caption{UML for the embarrassingly parallel framework}
\label{fig:uml}
\end{figure}
The only requirements for a program to be wrapped by framework are that it 
process the input in an embarrassingly parallel manner and that the input comes 
from a file. The processed results are must also written to a file. The 
framework supports two forms of input: a directory of files that contain that 
each contain one unit of work, which is represented by the Chunk class, or a 
single file that the framework can break up into Chunks. For example, the Ka/Ks 
Calculator's input consists of a header and then two protein sequence lines 
followed by a blank line. The framework uses that blank line to know when one 
Chunk of work has ended and another has begun. The user can input a regular 
expression to define when lines of input are seperated into Chunks. This occurs 
in the ChunkManager class.

The other purpose of the ChunkManager class is to combine the resulting Chunks 
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. The
ChunkManager's second job is to retrieve those results and construct a real
output file. The real output file must look the same as the output file would
look if the program were run serially; the output must come in the same order as
the inputs in the input file and headers may need to be deleted from the
separate files and added to the combined file. There are two other options if
the end-user does not want this behavior. They can choose to write their own
merger in Java, using the CustomMerge class, or write their own script to do the
merging.

@@

Lastly, the worker threads will keep records during execution. The threads 
records their uptime and the number of pieces of work they execute. The runtime 
for each piece of work is also recorded. This information will be used by us to
evaluate the effectiveness of our approach, but can also provide useful
information the can help a user tine a particular application on a particular
type of input data. The statistics for each thread can be found in the Worker 
subclass and the individual Chunk objects will store their own statstics about 
runtime and size.

@@

To observe how the framework changes the runtime of sequential programs when
using varying numbers of threads, we designed a simple experiment. The Ka/Ks 
Calculator was run through the framework with ten different samples, each 
containing sixteen units of work. Each sample was run by the framework at five 
different level of numbers of worker threads, starting at 1 and increasing by 
powers of two until the 100\% CPU usage (16 processors). These tests were run
with minimal other activity on the server.

@@

%TODO: Rewrite with new next steps
There are several next steps in the pipeline. In addition to using the
statistics to try to boost speed, adding savable configurations is at the top of
the list. This will allow users to use the same configuration without having to
go through the wizard if they need to redo a run. More extensive unit testing
will be implemented. The ability to chunk multiple pieces of work may be
implemented in order to avoid potential start up costs in the executable. 
Finally, other pleasing parallel problems will be executed by the wrapper so 
that more potential features will be discovered.

%TODO: Does this still make sense
Using Java was a good choice for implementation ease and future implementation
extension by others, but it may not produce the fastest results. Languages
like Hadoop, Clojure or other multithread oriented language may have been better
due to their parallel nature. In addition to a single machine working to process
the data quickly, we can also look to the cloud for help by using a networked 
solution like MPI or a cloud solution like Amazon instances. This would
potentially allow for an even greater time decrease because the number of
machines that could help is most likely larger than the number of processors on
one machine.

\begin{thebibliography}{1}
\bibitem{history}
Moler, Cleve. ``\emph{Matrix Computation on Distributed Memory Multiprocessors."
}, Hypercube Multiprocessors, 1986: Proceedings of the First Conference on
Hypercube Multiprocessors, Knoxville, Tennessee, August 24-27, 1985. By Michael
T. Heath. Philadelphia: SIAM, 1986. N. pag. Print.
\bibitem{kaks}
Zhang Z, Li J, Zhao XQ, Wang J, Wong GK, Yu J., \emph{Ka/Ks Calculator: 
calculating Ka and Ks through model selection and model averaging},
Genomics Proteomics Bioinformatics, November 2006.
\bibitem{cluster}
Rua, Xu, Wunsch, Donald II, \emph{Survey of Clustering Algorithms},
IEEE Transactions on Neural Networks VOL. 16, NO. 3, MAY 2005
\end{thebibliography}
